Sparse spectral and $p$-finite element methods on disk-slices and trapeziums I: zero boundary conditions

### 2019-04-04 papers/halfdisk/halfdisk.tex
Added new figures. Replaced transposes using letter T with top

\newcommand{\Dlt}{D^T_l}

@@

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pab}{P^{(a,b)}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

@@

$J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

\begin{align}
\Dnt := \begin{pmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{pmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1,\quad  \eta_0 = \frac{1}{\betaab_{n,n,6}}, \quad \xi = -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2.
\end{align}
It follows that we can apply $D_n^T$ in $O(n)$ complexity, and thereby construct $\bigPab_{n}(x,y)$ in optimal $O(n^2)$ complexity.

@@

There exist sparse operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= \bigW^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.

Further, we note that there exist "parameter transformation" Banded-Block-Banded matrix operators that increment/decrement the parameters, transforming the OPs from one (weighted or non-weighted) parameter space to another. We denote these operators as $T^{(a,b)\to(a+1,b)}$, $T^{(a,b)\to(a,b+1)}$ and $T^{(a,b)\to(a+1,b+1)}$ for conversion between non-weighted spaces, and similarly $T_W^{(a,b)\to(a-1,b)}$, $T_W^{(a,b)\to(a,b-1)}$ and $T_W^{(a,b)\to(a-1,b-1)}$ for conversion between weighted spaces, where:
\begin{align}
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a+1,b)} \Big)^T \: \bigP^{(a+1,b)}(x,y) \\
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^T \: \bigP^{(a,b+1)}(x,y) \\
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a+1,b+1)} \Big)^T \: \bigP^{(a+1,b+1)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a+1,b)} \Big)^T \: \bigW^{(a-1,b)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a,b+1)} \Big)^T \: \bigW^{(a,b-1)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a+1,b+1)} \Big)^T \: \bigW^{(a-1,b-1)}(x,y) \\
\end{align}

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{sparsityoflaplacian}
    	\caption{A "spy" plot of the $\laplacewii$ matrix operator for the Laplacian showing its sparsity.}
        \label{fig:sparsity}
	\centering
\end{figure}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bigP_n(x,y) &:= \begin{pmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{pmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align*}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T$, yielding an operator matrix of the same dimension as the input Jacobi matries. We can denote the resulting operator acting on coefficients in the $\bigPoo$ space by $V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T)$. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in FIGURE with $v(x,y) = xy^2$ as an example. In FIGURE we see the solution to the inhomogenous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the half-disk $\Omega$, with  $v(x,y) = xy^2$ and $f(x,y) = xy e^x$.

@@

Define $\Lambda^{(a,b)} := \langle \bigP^{(a,b)}, \: {\bigP^{(a,b)}}^T \rangle_{W^{(a,b)}}$ where $W^{(a,b)}$ is the weight with which the OPs in $\bigP^{(a,b)}$ are orthogonal with respect to. Note that due to orthogonality this is a diagonal matrix. We can choose a basis for $V_h$ by using the weighted orthogonal polynomials on $\element$ with parameters $a = b = 1$:
\begin{align}
\bigWii(x,y) &:= \begin{pmatrix}
		\bigWii_0(x,y) \\
		\bigWii_1(x,y) \\
		\bigWii_2(x,y) \\
		\vdots \\
		\bigWii_p(x,y)
	\end{pmatrix}, \\
\bigWii_n(x,y) &:= \begin{pmatrix}
		\Wii(x,y) \: \Pii_{n,0}(x,y) \\
		\vdots \\
		\Wii(a,y) \: \Pii_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
and rewrite (\ref{eqn:FEMweakform}) in matrix form:
\begin{align}
	a(u_h,v_h) &= \int_\element \nabla u_h \cdot \nabla v_h \: d\mathbf{x} \\
	&= \int_\element \begin{pmatrix}
					\partial_x v_h \\
					\partial_y v_h
				\end{pmatrix}^T 
				\begin{pmatrix}
					\partial_x u_h \\
					\partial_y u_h
				\end{pmatrix}
				\\
	&= \int_\element \begin{pmatrix}
					\bigPoo^T \Wii_x \mathbf{v} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{v}
				\end{pmatrix}^T 
				\begin{pmatrix}
					\bigPoo^T \Wii_x \mathbf{u} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}
				\end{pmatrix}
				\\
	&= \int_\element \Big( \mathbf{v}^T {\Wii_x}^T \bigPoo \bigPoo^T \Wii_x \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^T ({T^{(1,0)\to(0,0)} \Wii_y})^T \bigPoo \bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}  \Big) \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \Big( {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y \Big) \: \mathbf{u}
\end{align}
where $\mathbf{u}, \mathbf{v}$ are the coefficient vectors of the expansions of $u_h, v_h \in V_h$ respectively in the $V_h$ basis ($\bigWii$ OPs), and
\begin{align}
	L(v_h) &= \int_\element \: v_h \: f \: d\mathbf{x} \\
	&= \int_\element \: \mathbf{v}^T \: \bigWii \: \bigPii^T \: \mathbf{f} \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \langle \bigPii, {\bigPii}^T \rangle_{\Wii} \: d\mathbf{x} \\
	&= \mathbf{v}^T \Lambda^{(1,1)} \: \mathbf{f},
\end{align}
where $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y)$ in the $\bigPii$ OP basis.

Since (\ref{eqn:FEMweakform}) is equivalent to stating that
\begin{align}
	L(\Wii \Pii_{n,k}) = a(u_h,\Wii \Pii_{n,k}) \quad \forall \: n = 0,\dots,p, \: k = 0,\dots,n,
\end{align}
(i.e. holds for all basis functions of $V_h$) by choosing $v_h$ as each basis function, we can equivalently write the linear system for our finite element problem as:
\begin{align}
A\mathbf{u} = \tilde{\mathbf{f}}.
\end{align}
where the (element) stiffness matrix $A$ is defined by 
\begin{align}
A = {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y, 
\end{align}
and the load vector $\tilde{\mathbf{f}}$ is given by 
\begin{align}
\tilde{\mathbf{f}} = \Lambda^{(1,1)} \: \mathbf{f}.
\end{align}

>>>>

\newcommand{\Dlt}{D^\top_l}

@@

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pab}{P^{(a,b)}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

@@

$J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

\begin{align}
\Dnt := \begin{pmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{pmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1,\quad  \eta_0 = \frac{1}{\betaab_{n,n,6}}, \quad \xi = -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2.
\end{align}
It follows that we can apply $\Dnt$ in $O(n)$ complexity, and thereby construct $\bigPab_{n}(x,y)$ in optimal $O(n^2)$ complexity.

@@

There exist sparse operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^\top \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^\top \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1)}(x,y)^\top \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= \bigW^{(a,b-1)}(x,y)^\top \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.

Further, we note that there exist "parameter transformation" Banded-Block-Banded matrix operators that increment/decrement the parameters, transforming the OPs from one (weighted or non-weighted) parameter space to another. We denote these operators as $T^{(a,b)\to(a+1,b)}$, $T^{(a,b)\to(a,b+1)}$ and $T^{(a,b)\to(a+1,b+1)}$ for conversion between non-weighted spaces, and similarly $T_W^{(a,b)\to(a-1,b)}$, $T_W^{(a,b)\to(a,b-1)}$ and $T_W^{(a,b)\to(a-1,b-1)}$ for conversion between weighted spaces, where:
\begin{align}
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a+1,b)} \Big)^\top \: \bigP^{(a+1,b)}(x,y) \\
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^\top \: \bigP^{(a,b+1)}(x,y) \\
\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a+1,b+1)} \Big)^\top \: \bigP^{(a+1,b+1)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a+1,b)} \Big)^\top \: \bigW^{(a-1,b)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a,b+1)} \Big)^\top \: \bigW^{(a,b-1)}(x,y) \\
\bigW^{(a,b)}(x,y) &= \Big(T_W^{(a,b)\to(a+1,b+1)} \Big)^\top \: \bigW^{(a-1,b-1)}(x,y) \\
\end{align}

\begin{figure}
	\begin{subfigure}[t]{0.45\textwidth}
	\includegraphics[scale=0.4]{sparsityoflaplacian-w11}
	\end{subfigure}
	\begin{subfigure}[t]{0.45\textwidth}
	\includegraphics[scale=0.4]{sparsityoflaplacian-w22}
	\end{subfigure}
	\begin{subfigure}[t]{0.45\textwidth}
	\includegraphics[scale=0.4]{sparsityoflaplacian-00}
	\end{subfigure}
	\begin{subfigure}[t]{0.45\textwidth}
	\includegraphics[scale=0.4]{sparsityofhelmholtz}
	\end{subfigure}
	\centering
    	\caption{"Spy" plots of the different operator matrices for the Laplacian operator $\Delta$, as well as the operator matrix for the Helmholtz equation, showing their sparsity. Top Left: $\laplacewii$. Top Right: $\laplacewtt$. Bottom Left: $\laplaceoo$. Bottom Right: the weighted variable coefficient Helmholtz operator $\laplacewii + k^2 \: T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}$ for $v(x,y) = xy^2$ and $k = 0.25$.}
        \label{fig:sparsity}
\end{figure}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^\top \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bigP_n(x,y) &:= \begin{pmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{pmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align*}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top$, yielding an operator matrix of the same dimension as the input Jacobi matries. We can denote the resulting operator acting on coefficients in the $\bigPoo$ space by $V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top)$. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in FIGURE with $v(x,y) = xy^2$ as an example. In FIGURE we see the solution to the inhomogenous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the half-disk $\Omega$, with  $v(x,y) = xy^2$ and $f(x,y) = xy e^x$.

@@

Define $\Lambda^{(a,b)} := \langle \bigP^{(a,b)}, \: {\bigP^{(a,b)}}^\top \rangle_{W^{(a,b)}}$ where $W^{(a,b)}$ is the weight with which the OPs in $\bigP^{(a,b)}$ are orthogonal with respect to. Note that due to orthogonality this is a diagonal matrix. We can choose a basis for $V_h$ by using the weighted orthogonal polynomials on $\element$ with parameters $a = b = 1$:
\begin{align}
\bigWii(x,y) &:= \begin{pmatrix}
		\bigWii_0(x,y) \\
		\bigWii_1(x,y) \\
		\bigWii_2(x,y) \\
		\vdots \\
		\bigWii_p(x,y)
	\end{pmatrix}, \\
\bigWii_n(x,y) &:= \begin{pmatrix}
		\Wii(x,y) \: \Pii_{n,0}(x,y) \\
		\vdots \\
		\Wii(a,y) \: \Pii_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
and rewrite (\ref{eqn:FEMweakform}) in matrix form:
\begin{align}
	a(u_h,v_h) &= \int_\element \nabla u_h \cdot \nabla v_h \: d\mathbf{x} \\
	&= \int_\element \begin{pmatrix}
					\partial_x v_h \\
					\partial_y v_h
				\end{pmatrix}^\top 
				\begin{pmatrix}
					\partial_x u_h \\
					\partial_y u_h
				\end{pmatrix}
				\\
	&= \int_\element \begin{pmatrix}
					\bigPoo^\top \Wii_x \mathbf{v} \\
					\bigPoo^\top T^{(1,0)\to(0,0)} \Wii_y \mathbf{v}
				\end{pmatrix}^\top 
				\begin{pmatrix}
					\bigPoo^\top \Wii_x \mathbf{u} \\
					\bigPoo^\top T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}
				\end{pmatrix}
				\\
	&= \int_\element \Big( \mathbf{v}^\top {\Wii_x}^\top \bigPoo \bigPoo^\top \Wii_x \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^\top ({T^{(1,0)\to(0,0)} \Wii_y})^\top \bigPoo \bigPoo^\top T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}  \Big) \: d\mathbf{x} \\
	&= \mathbf{v}^\top \: \Big( {\Wii_x}^\top \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^\top \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y \Big) \: \mathbf{u}
\end{align}
where $\mathbf{u}, \mathbf{v}$ are the coefficient vectors of the expansions of $u_h, v_h \in V_h$ respectively in the $V_h$ basis ($\bigWii$ OPs), and
\begin{align}
	L(v_h) &= \int_\element \: v_h \: f \: d\mathbf{x} \\
	&= \int_\element \: \mathbf{v}^\top \: \bigWii \: \bigPii^\top \: \mathbf{f} \: d\mathbf{x} \\
	&= \mathbf{v}^\top \: \langle \bigPii, {\bigPii}^\top \rangle_{\Wii} \: d\mathbf{x} \\
	&= \mathbf{v}^\top \Lambda^{(1,1)} \: \mathbf{f},
\end{align}
where $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y)$ in the $\bigPii$ OP basis.

Since (\ref{eqn:FEMweakform}) is equivalent to stating that
\begin{align}
	L(\Wii \Pii_{n,k}) = a(u_h,\Wii \Pii_{n,k}) \quad \forall \: n = 0,\dots,p, \: k = 0,\dots,n,
\end{align}
(i.e. holds for all basis functions of $V_h$) by choosing $v_h$ as each basis function, we can equivalently write the linear system for our finite element problem as:
\begin{align}
A\mathbf{u} = \tilde{\mathbf{f}}.
\end{align}
where the (element) stiffness matrix $A$ is defined by 
\begin{align}
A = {\Wii_x}^\top \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^\top \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y, 
\end{align}
and the load vector $\tilde{\mathbf{f}}$ is given by 
\begin{align}
\tilde{\mathbf{f}} = \Lambda^{(1,1)} \: \mathbf{f}.
\end{align}
